**Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces**

### 2020-11-16 submission/manuscript.Rmd
Coauthors review done.

\newpage
\linenumbers
\sisetup{mode=text}
\setlength\parindent{24pt}

@@

All these ecological roles are carried out by a taxonomically diverse community of microorganisms. At higher taxonomic ranks a core set of macrophyte epiphytic taxa was described consisting mainly of *Alphaproteobacteria*, *Gammaproteobacteria*, *Desulfobacterota*, *Bacteroidota*, *Cyanobacteria*, *Actinobacteriota*, *Firmicutes*, *Planctomycetota*, *Chloroflexi* and *Verrucomicrobiota* [@Lachnit2011; @Crump2008; @Tujula2010]. In contrast, at lower taxonomic ranks host specific microbial communities were found [@Roth-Schulze2016; @Lachnit2011]. Recently, it was shown that even different morphological niches within the same alga had a higher influence on the composition of the bacterial community than biogeography or environmental factors [@Morrissey2019]. While the microbial community composition varies between host species, metagenomic analyses revealed that the majority of the microbial functions are conserved [@Burke2011; @Roth-Schulze2016]. This discrepancy between the microbial taxonomic and functional composition might be explained by the lottery hypothesis. It postulates that an initial random colonization step takes places from a set of functionally equivalent taxonomic groups resulting in taxonomically different epiphytic communities sharing a core set of functional genes [@Burke2011; @Roth-Schulze2016]. In addition, some of the variation in the reported data could be attributed to different techniques used in these studies, such as different protocols for epiphytic cell detachment and/or DNA isolation, as no standard protocol has been yet established to study epiphytic communities [@Korlevica, submitted; @Ugarelli2019].

@@

Sequencing of the 16S rRNA V4 region yielded a total of `r million_sequences` million sequences after quality curation and exclusion of sequences without known relatives (no relative sequences) and eukaryotic, chloroplast and mitochondrial sequences (\autoref{supp-nseq_notus}). A total of `r samples_epiphytes` samples originating from epiphytic archaeal and bacterial communities associated with surfaces of the seagrass *C. nodosa* and the macroalga *C. cylindracea* were analyzed. In addition, `r samples_seawater` samples (one of the samples was sequenced twice) originating from the ambient seawater were also processed for comparison. The number of reads per sample ranged between `r min_max$min` and `r min_max$max` sequences (\autoref{supp-nseq_notus}). Even when the highest sequencing effort was applied the rarefaction curves did not level off as commonly observed in high-throughput 16S rRNA amplicon sequencing approaches (\autoref{supp-rarefaction}). Following quality curation and exclusion of sequences as mentioned above reads were clustered into `r no_otus` different OTUs at a similarity level of 97 \si{\percent}. Read numbers were normalized to the minimum number of sequences (`r min_max$min`, \autoref{supp-nseq_notus}) through rarefaction resulting in `r no_otus_rarefied` different OTUs with `r min(otu_min_max)` to `r max(otu_min_max)` OTUs per sample (\autoref{supp-calculators}). To determine seasonal changes in richness and diversity the observed number of OTUs, Chao1, ACE, Exponential Shannon [@Jost2006] and Inverse Simpson were calculated after normalization through rarefaction. Generally, richness estimators and diversity indices showed similar trends. On average, higher values were found for *C. cylindracea* (mixed [Number of OTUs, `r S.obs_fcam$mean` ± `r S.obs_fcam$sd` OTUs] and monospecific [Number of OTUs, `r S.obs_fca$mean` ± `r S.obs_fca$sd` OTUs]) than for *C. nodosa* (Number of OTUs, `r S.obs_fcym$mean` ± `r S.obs_fcym$sd` OTUs) and lowest values were obtained for the microbial community of the ambient seawater (Number of OTUs, `r S.obs_f$mean` ± `r S.obs_f$sd` OTUs) (\autoref{supp-calculators}). Seasonal changes did not reveal such large dissimilarities. *C. nodosa* communities showed a slow increase towards the end of the study, while *C. cylindracea* (mixed and monospecific) communities were characterized by slightly higher values in spring and summer than in autumn and winter (\autoref{supp-calculators}).

@@

# Calculating matrices of similarity indices
jaccard <- vegdist(rarefied_metadata, method="jaccard", binary=T)
jaccard <- as_tibble(data.frame(t(combn(rownames(rarefied_metadata), 2)), as.numeric(jaccard)),
                     .name_repair= ~c("V1", "V2", "jaccard"))

bray <- vegdist(rarefied_metadata, method="bray", binary=F)
bray <- as_tibble(data.frame(t(combn(rownames(rarefied_metadata), 2)), as.numeric(bray)),
                  .name_repair= ~c("V1", "V2", "bray"))

@@

# Selection of groups for plotting
select <- filter(community,
               taxlevel==6 &
               str_detect(rankID, filter(community, str_detect(taxon, "^Cyanobacteria$"))$rankID)) %>%
  filter_at(6:ncol(.), any_vars(. >= 1))

plot <- filter(community,
               taxlevel==6 &
               str_detect(rankID, filter(community, str_detect(taxon, "^Cyanobacteria$"))$rankID)) %>%
  mutate_at(5:ncol(.), list(~. / sum(.) * 100)) %>%
  ungroup() %>%
  filter(taxon %in% select$taxon) %>%
  bind_rows(summarise_all(., list(~ifelse(is.numeric(.), 100-sum(.), paste("Other_Cyanobacteria")))))

@@

```{r alpha}
# Selection of groups for plotting
select <- filter(community,
               taxlevel==6 &
               str_detect(rankID, filter(community, str_detect(taxon, "^Alphaproteobacteria$"))$rankID)) %>%
  filter_at(6:ncol(.), any_vars(. >= 2))

plot <- filter(community,
               taxlevel==6 &
               str_detect(rankID, filter(community, str_detect(taxon, "^Alphaproteobacteria$"))$rankID)) %>%
  mutate_at(5:ncol(.), list(~. / sum(.) * 100)) %>%
  ungroup() %>%
  filter(rankID %in% select$rankID) %>%
  bind_rows(summarise_all(., list(~ifelse(is.numeric(.), 100-sum(.), paste("Other_Alphaproteobacteria"))))) %>%
# Remove last digit from rankID so for the next step
  mutate(rankID=if_else(taxon=="uncultured", str_replace(rankID, "\\.\\d+$", ""), rankID))

@@

\newpage
## Discussion
Surfaces of marine macrophytes harbor biofilms consisting of diverse microbial taxa [@Egan2013; @Tarquinio2019]. No standard protocol has been developed to study these macophyte-associated microbes [@Ugarelli2019]. Different procedures for removal of microbial cells from host surfaces are described, such as host tissue shaking [@Noges2010], scraping [@Uku2007], swabbing [@Mancuso2016] and ultrasonication [@Cai2014]. All these methods result in different removal efficiencies but none was enabling a complete removal of attached microbial cells based on our experience. In the present study, we applied a removal protocol [@Korlevica, submitted] based on direct cellular lysis [@Burke2009]. The application of a direct lysis procedure coupled with a high sampling frequency and Illumina amplicon sequencing has enabled us to described in detail the bacterial and archaeal communities associated with the surfaces of two marine macrophytes, *C. nodosa* and *C. cylindracea*.

@@

\newpage
## Experimental procedures
### Sampling
Sampling was performed in the Bay of Funtana, northern Adriatic Sea (\ang{45;10;39} N, \ang{13;35;42} E). Leaves and thalli were sampled in a meadow of *Cymodocea nodosa* invaded by the invasive *Caulerpa cylindracea* (mixed settlement) and in a monospecific settlement of *Caulerpa cylindracea* located in the proximity of the meadow at approximately monthly intervals from December 2017 to October 2018 (\autoref{supp-nseq_notus}). Leaves and thalli were collected by diving and transported to the laboratory in containers placed on ice and filled with seawater collected at the sampling site. Upon arrival to the laboratory, *C. nodosa* leaves were cut into sections of 1 -- 2 \si{\cm}, while *C. cylindracea* thalli were cut into 5 -- 8 \si{\cm} long sections. Leaves and thalli were washed three times with sterile artificial seawater (ASW) to remove loosely attached microbial cells. Ambient seawater was collected in 10 \si{\l} containers by diving and transported to the laboratory where the whole container volume was filtered through a 20 \si{\um} net. The filtrate was further sequentially filtered through 3 \si{\um} and 0.2 \si{\um} polycarbonate membrane filters (Whatman, United Kingdom) using a peristaltic pump. Filters were briefly dried at room temperature and stored at \num{-80} \si{\degreeCelsius}. Seawater samples were also collected approximately monthly from July 2017 to October 2018.

### DNA isolation
DNA from surfaces of *C. nodosa* and *C. cylindracea* was isolated using a previously modified and adapted protocol that allows for a selective epiphytic DNA isolation [@Korlevica, submitted; @Massana1997]. Briefly, leaves and thalli are incubated in a lysis buffer and treated with lysozyme and proteinase K. Following the incubations, the mixture containing lysed epiphytic cells was separated from the leaves and thalli and extracted using phenol-chloroform. Finally, the extracted DNA was precipitated using isopropanol. DNA from seawater picoplankton was extracted from 0.2 \si{\um} polycarbonate filters according to @Massana1997 with a slight modification. Following the phenol-chloroform extraction, 1/10 of chilled 3 \si{\Molar} sodium acetate (pH 5.2) was added. DNA was precipitated by adding 1 volume of chilled isopropanol, incubating the mixtures overnight at \num{-20} \si{\degreeCelsius} and centrifuging at 20,000 × *g* and 4 \si{\degreeCelsius} for 21 \si{\minute}. The pellet was washed twice with 500 \si{\ul} of chilled 70 \si{\percent} ethanol and centrifuged after each washing step at 20,000 × *g* and 4 \si{\degreeCelsius} for 5 \si{\minute}. Dried pellets were re-suspended in 50 -- 100 \si{\ul} of deionized water.

### Illumina 16S rRNA sequencing
Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was performed as described previously [@Korlevica, submitted]. The V4 region of the 16S rRNA gene was amplified using a two-step PCR procedure. In the first PCR, the 515F (5'-GTGYCAGCMGCCGCGGTAA-3') and 806R (5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome Project (http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/) were used [@Caporaso2012; @Apprill2015; @Parada2016]. These primers contained on their 5' end a tagged sequence. Purified PCR products were sent for Illumina MiSeq sequencing at IMGM Laboratories, Martinsried, Germany. Prior to sequencing at IMGM, the second PCR amplification of the two-step PCR procedure was performed using primers targeting the tagged region incorporated in the first PCR. In addition, these primers contained adapter and sample-specific index sequences. Beside samples, a positive and negative control for each sequencing batch was sequenced. The negative control comprised PCR reactions without DNA template, while for a positive control a mock community composed of evenly mixed DNA material originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Sequences obtained in this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB37267.

@@

## Acknowledgments
This work was founded by the Croatian Science Foundation through the MICRO-SEAGRASS project (IP-2016-06-7118). We would like to thank Margareta Buterer for technical support, Paolo Paliaga for help during sampling and the University Computing Center of the University of Zagreb for access to the computer cluster Isabella.

@@

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant (> 1 \\si{\\percent}) bacterial and archaeal sequences on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{community}", fig.pos="H"}
include_graphics("../results/figures/community_bar_plot.jpg")
```

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant (> 1 \\si{\\percent}) cyanobacterial sequences on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of cyanobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{cyano}", fig.pos="H"}
include_graphics("../results/figures/cyanobacteria_bar_plot.jpg")
```

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant (> 2 \\si{\\percent}) sequences within the \\textit{Bacteroidota} on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of sequences classified as \\textit{Bacteroidota} in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{bactero}", fig.pos="H"}
include_graphics("../results/figures/bacteroidota_bar_plot.jpg")
```

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant (> 2 \\si{\\percent}) alphaproteobacterial sequences on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of alphaproteobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{alpha}", fig.pos="H"}
include_graphics("../results/figures/alphaproteobacteria_bar_plot.jpg")
```

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant (> 2 \\si{\\percent}) gammaproteobacterial sequences on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of gammaproteobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{gamma}", fig.pos="H"}
include_graphics("../results/figures/gammaproteobacteria_bar_plot.jpg")

@@

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant (> 1 \\si{\\percent}) sequences within the \\textit{Desulfobacterota} on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of sequences classified as \\textit{Desulfobacterota} in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{desulfo}", fig.pos="H"}
include_graphics("../results/figures/desulfobacterota_bar_plot.jpg")
```

>>>>

\newpage
[]: # (\linenumbers)
\sisetup{mode=text}
\setlength\parindent{24pt}

@@

All these ecological roles are carried out by a taxonomically diverse community of microorganisms. At higher taxonomic ranks a core set of macrophyte epiphytic taxa was described consisting mainly of *Alphaproteobacteria*, *Gammaproteobacteria*, *Desulfobacterota*, *Bacteroidota*, *Cyanobacteria*, *Actinobacteriota*, *Firmicutes*, *Planctomycetota*, *Chloroflexi* and *Verrucomicrobiota* [@Lachnit2011; @Crump2008; @Tujula2010]. In contrast, at lower taxonomic ranks host specific microbial communities were found [@Roth-Schulze2016; @Lachnit2011]. Recently, it was shown that even different morphological niches within the same alga had a higher influence on the composition of the bacterial community than biogeography or environmental factors [@Morrissey2019]. While the microbial community composition varies between host species, metagenomic analyses revealed that the majority of the microbial functions are conserved [@Burke2011; @Roth-Schulze2016]. This discrepancy between the microbial taxonomic and functional composition might be explained by the lottery hypothesis. It postulates that an initial random colonization step takes places from a set of functionally equivalent taxonomic groups resulting in taxonomically different epiphytic communities sharing a core set of functional genes [@Burke2011; @Roth-Schulze2016]. In addition, some of the variation in the reported data could be attributed to different techniques used in these studies, such as different protocols for epiphytic cell detachment and/or DNA isolation, as no standard protocol has been yet established to study epiphytic communities [@Korlevic, submitted; @Ugarelli2019].

@@

Sequencing of the 16S rRNA V4 region yielded a total of `r million_sequences` million sequences after quality curation and exclusion of sequences without known relatives (no relative sequences) and eukaryotic, chloroplast and mitochondrial sequences (\autoref{supp-nseq_notus}). A total of `r samples_epiphytes` samples originating from epiphytic archaeal and bacterial communities associated with surfaces of the seagrass *C. nodosa* and the macroalga *C. cylindracea* were analyzed. In addition, `r samples_seawater` samples (one of the samples was sequenced twice) originating from the ambient seawater were also processed for comparison. The number of reads per sample ranged between `r min_max$min` and `r min_max$max` sequences (\autoref{supp-nseq_notus}). Even when the highest sequencing effort was applied the rarefaction curves did not level off as commonly observed in high-throughput 16S rRNA amplicon sequencing approaches (\autoref{supp-rarefaction}). Following quality curation and exclusion of sequences as mentioned above reads were clustered into `r no_otus` different OTUs at a similarity level of 97 \si{\percent}. Read numbers were normalized to the minimum number of sequences (`r min_max$min`, \autoref{supp-nseq_notus}) through rarefaction resulting in `r no_otus_rarefied` different OTUs with `r min(otu_min_max)` to `r max(otu_min_max)` OTUs per sample (\autoref{supp-calculators}). To determine seasonal changes in richness and diversity the observed number of OTUs, Chao1, ACE, Exponential Shannon and Inverse Simpson [@Jost2006] were calculated after normalization through rarefaction. Generally, richness estimators and diversity indices showed similar trends. On average, higher values were found for *C. cylindracea* (mixed [Number of OTUs, `r S.obs_fcam$mean` ± `r S.obs_fcam$sd` OTUs] and monospecific [Number of OTUs, `r S.obs_fca$mean` ± `r S.obs_fca$sd` OTUs]) than for *C. nodosa* (Number of OTUs, `r S.obs_fcym$mean` ± `r S.obs_fcym$sd` OTUs) and lowest values were obtained for the microbial community of the ambient seawater (Number of OTUs, `r S.obs_f$mean` ± `r S.obs_f$sd` OTUs) (\autoref{supp-calculators}). Seasonal changes did not reveal such large dissimilarities. *C. nodosa* communities showed a slow increase towards the end of the study, while *C. cylindracea* (mixed and monospecific) communities were characterized by slightly higher values in spring and summer than in autumn and winter (\autoref{supp-calculators}).

@@

# Calculating matrices of similarity indices
jaccard <- as.matrix(vegdist(rarefied_metadata, method="jaccard", binary=T))
jaccard[lower.tri(jaccard, diag=TRUE)] <- NA
jaccard <- as_tibble(jaccard) %>%
  add_column("V1"=rownames(jaccard), .before=TRUE) %>%
  gather(key="V2", value="jaccard", 2:ncol(.)) %>%
  filter(!is.na(jaccard))

bray <- as.matrix(vegdist(rarefied_metadata, method="bray", binary=F))
bray[lower.tri(bray, diag=TRUE)] <- NA
bray <- as_tibble(bray) %>%
  add_column("V1"=rownames(bray), .before=TRUE) %>%
  gather(key="V2", value="bray", 2:ncol(.)) %>%
  filter(!is.na(bray))

@@

# Selection of groups for plotting
select <- filter(community,
               taxlevel==6 &
               str_detect(rankID, paste0(filter(community, str_detect(taxon, "^Cyanobacteria$"))$rankID, "\\."))) %>%
  filter_at(6:ncol(.), any_vars(. >= 1))

plot <- filter(community,
               taxlevel==6 &
               str_detect(rankID, paste0(filter(community, str_detect(taxon, "^Cyanobacteria$"))$rankID, "\\."))) %>%
  mutate_at(5:ncol(.), list(~. / sum(.) * 100)) %>%
  ungroup() %>%
  filter(taxon %in% select$taxon) %>%
  bind_rows(summarise_all(., list(~ifelse(is.numeric(.), 100-sum(.), paste("Other_Cyanobacteria")))))

@@

```{r alpha}
# Selection of groups for plotting
select <- filter(community,
               taxlevel==6 &
               str_detect(rankID, paste0(filter(community, str_detect(taxon, "^Alphaproteobacteria$"))$rankID, "\\."))) %>%
  filter_at(6:ncol(.), any_vars(. >= 2))

plot <- filter(community,
               taxlevel==6 &
               str_detect(rankID, paste0(filter(community, str_detect(taxon, "^Alphaproteobacteria$"))$rankID, "\\."))) %>%
  mutate_at(5:ncol(.), list(~. / sum(.) * 100)) %>%
  ungroup() %>%
  filter(rankID %in% select$rankID) %>%
  bind_rows(summarise_all(., list(~ifelse(is.numeric(.), 100-sum(.), paste("Other_Alphaproteobacteria"))))) %>%
# Remove last digit from rankID so for the next step
  mutate(rankID=if_else(taxon=="uncultured", str_replace(rankID, "\\.\\d+$", ""), rankID))

@@

\newpage
## Discussion
Surfaces of marine macrophytes harbor biofilms consisting of diverse microbial taxa [@Egan2013; @Tarquinio2019]. No standard protocol has been developed to study these macophyte-associated microbes [@Ugarelli2019]. Different procedures for removal of microbial cells from host surfaces are described, such as host tissue shaking [@Noges2010], scraping [@Uku2007], swabbing [@Mancuso2016] and ultrasonication [@Cai2014]. All these methods result in different removal efficiencies but none was enabling a complete removal of attached microbial cells based on our experience. In the present study, we applied a removal protocol [@Korlevic, submitted] based on direct cellular lysis [@Burke2009]. The application of a direct lysis procedure coupled with a high sampling frequency and Illumina amplicon sequencing has enabled us to described in detail the bacterial and archaeal communities associated with the surfaces of two marine macrophytes, *C. nodosa* and *C. cylindracea*.

@@

\newpage
## Experimental procedures
### Sampling
Sampling was performed in the Bay of Funtana, northern Adriatic Sea (\ang{45;10;39} N, \ang{13;35;42} E). Leaves and thalli were sampled in a meadow of *Cymodocea nodosa* invaded by the invasive *Caulerpa cylindracea* (mixed settlement) and in a monospecific settlement of *Caulerpa cylindracea* located in the proximity of the meadow at approximately monthly intervals from December 2017 to October 2018 (\autoref{supp-nseq_notus}). Leaves and thalli were collected by diving and transported to the laboratory in containers placed on ice and filled with seawater collected at the sampling site. Upon arrival to the laboratory, *C. nodosa* leaves were cut into sections of 1 -- 2 \si{\cm}, while *C. cylindracea* thalli were cut into 5 -- 8 \si{\cm} long sections. Leaves and thalli were washed three times with sterile artificial seawater (ASW) to remove loosely attached microbial cells. Ambient seawater was collected in 10 \si{\l} containers by diving and transported to the laboratory where 10 -- 20 \si{\l} were filtered through a 20 \si{\um} net. The filtrate was further sequentially filtered through 3 \si{\um} and 0.2 \si{\um} polycarbonate membrane filters (Whatman, United Kingdom) using a peristaltic pump. Filters were briefly dried at room temperature and stored at \num{-80} \si{\degreeCelsius}. Seawater samples were also collected approximately monthly from July 2017 to October 2018.

### DNA isolation
DNA from surfaces of *C. nodosa* and *C. cylindracea* was isolated using a previously modified and adapted protocol that allows for a selective epiphytic DNA isolation [@Korlevic, submitted; @Massana1997]. Briefly, leaves and thalli are incubated in a lysis buffer and treated with lysozyme and proteinase K. Following the incubations, the mixture containing lysed epiphytic cells was separated from the leaves and thalli and extracted using phenol-chloroform. Finally, the extracted DNA was precipitated using isopropanol. DNA from seawater picoplankton was extracted from 0.2 \si{\um} polycarbonate filters according to @Massana1997 with a slight modification. Following the phenol-chloroform extraction, 1/10 of chilled 3 \si{\Molar} sodium acetate (pH 5.2) was added. DNA was precipitated by adding 1 volume of chilled isopropanol, incubating the mixtures overnight at \num{-20} \si{\degreeCelsius} and centrifuging at 20,000 × *g* and 4 \si{\degreeCelsius} for 21 \si{\minute}. The pellet was washed twice with 500 \si{\ul} of chilled 70 \si{\percent} ethanol and centrifuged after each washing step at 20,000 × *g* and 4 \si{\degreeCelsius} for 5 \si{\minute}. Dried pellets were re-suspended in 50 -- 100 \si{\ul} of deionized water.

### Illumina 16S rRNA sequencing
Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was performed as described previously [@Korlevic, submitted]. The V4 region of the 16S rRNA gene was amplified using a two-step PCR procedure. In the first PCR, the 515F (5'-GTGYCAGCMGCCGCGGTAA-3') and 806R (5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome Project (http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/) were used [@Caporaso2012; @Apprill2015; @Parada2016]. These primers contained on their 5' end a tagged sequence. Purified PCR products were sent for Illumina MiSeq sequencing at IMGM Laboratories, Martinsried, Germany. Prior to sequencing at IMGM, the second PCR amplification of the two-step PCR procedure was performed using primers targeting the tagged region incorporated in the first PCR. In addition, these primers contained adapter and sample-specific index sequences. Beside samples, a positive and negative control for each sequencing batch was sequenced. The negative control comprised PCR reactions without DNA template, while for a positive control a mock community composed of evenly mixed DNA material originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Sequences obtained in this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB37267.

@@

## Acknowledgments
This work was founded by the Croatian Science Foundation through the MICRO-SEAGRASS project (project number IP-2016-06-7118). ZZ and GJH were supported by the Austrian Science Fund (FWF) project ARTEMIS (project number P28781-B21). We would like to thank Margareta Buterer for technical support, Paolo Paliaga for help during sampling and the University Computing Center of the University of Zagreb for access to the computer cluster Isabella.

@@

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant ($\\geq$ 1 \\si{\\percent}) bacterial and archaeal sequences on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{community}", fig.pos="H"}
include_graphics("../results/figures/community_bar_plot.jpg")
```

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant ($\\geq$ 1 \\si{\\percent}) cyanobacterial sequences on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of cyanobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{cyano}", fig.pos="H"}
include_graphics("../results/figures/cyanobacteria_bar_plot.jpg")
```

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant ($\\geq$ 2 \\si{\\percent}) sequences within the \\textit{Bacteroidota} on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of sequences classified as \\textit{Bacteroidota} in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{bactero}", fig.pos="H"}
include_graphics("../results/figures/bacteroidota_bar_plot.jpg")
```

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant ($\\geq$ 2 \\si{\\percent}) alphaproteobacterial sequences on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of alphaproteobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{alpha}", fig.pos="H"}
include_graphics("../results/figures/alphaproteobacteria_bar_plot.jpg")
```

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant ($\\geq$ 1 \\si{\\percent}) gammaproteobacterial sequences on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of gammaproteobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{gamma}", fig.pos="H"}
include_graphics("../results/figures/gammaproteobacteria_bar_plot.jpg")

@@

```{r out.width="85%", fig.align="center", fig.cap="Taxonomic classification and relative contribution of the most abundant ($\\geq$ 1 \\si{\\percent}) sequences within the \\textit{Desulfobacterota} on the surfaces of the macrophytes \\textit{C. nodosa} (mixed settlement) and \\textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of sequences classified as \\textit{Desulfobacterota} in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\\label{desulfo}", fig.pos="H"}
include_graphics("../results/figures/desulfobacterota_bar_plot.jpg")
```
